---
title: "Choroid Contamination: TTR Groups"
author: "Kennedi Todd"
date: "8/26/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setwd, include=FALSE}
knitr::opts_knit$set(root.dir = "~/TTR/03_GTEx/")
setwd("~/TTR/03_GTEx/")

```

Load packages.
```{r packages, results='hide'}
library(BiocParallel)
library(dplyr)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(gplots)
library(grid)
library(gridExtra)
library(knitr)
library(stringr)
library(variancePartition)
```

Read in files.
```{r}
setwd("~/TTR/03_GTEx/")

# read in counts
counts <- read.delim("Files/Brain_Hippocampus", 
                     header = TRUE, 
                     sep = "\t",
                     row.names = 1)
# read in metadata
metadata <- read.delim(
  "Files/GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt",
  header = TRUE,
  sep = "\t")
# read in gene annotation
gtf.file <- "Files/gencode.v26.GRCh38.genes.gtf"
gtf.gr <- rtracklayer::import(gtf.file)
gtf.df <- as.data.frame(gtf.gr)
gtf.df <- gtf.df[,c("type","gene_type","gene_name","gene_id","seqnames")]
gtf.genes <- gtf.df[gtf.df$type == "gene",]
```

Reformat metadata and counts.
```{r}
# reformat metadata
colnames(metadata) <- c("subject_id","sex","age","hardy_scale")

# reformat counts
rownames(counts) <- counts$Name
counts <- counts[gtf.genes$gene_id,] # put counts in same order as gtf.genes
all.equal(counts$Name, gtf.genes$gene_id) # check
counts <- counts[,-c(1,2,200,201)]  # remove mean, median, gene_id, gene_name
```

Rename counts columns to match metadata.
```{r}
# change all hyphens to periods
metadata$subject_id <- gsub("-",".",metadata$subject_id)
rownames(metadata) <- metadata$subject_id

# parse subject_id out of counts column names
old.names <- colnames(counts)  # GTEX.13PDP.0011.R1a.SM.5PNX5
new.names <- str_match(old.names, "(GTEX\\.[\\d\\w]+)\\.\\d+")[,2]
head(cbind(old.names, new.names)) # check
colnames(counts) <- new.names

# Put metadata rows in the same order as counts columns
metadata <- metadata[colnames(counts),]

# check
all.equal(rownames(metadata), colnames(counts))
```

Create dge object
```{r}
dge <- DGEList(counts, 
               samples = metadata, 
               genes = gtf.genes)
```

Look to see if any NA values or replicated gene names.
```{r}
table(is.na(gtf.genes$gene_name))
table(duplicated(gtf.genes$gene_name))
```
There are no NA values.  However, 1,608 gene names are repetitive.

View the repetitive gene names.
```{r}
table(gtf.genes$gene_name[duplicated(gtf.genes$gene_name)])
```

View the gene_type for the repetitive genes.
This will help determine if we should remove repetitive gene names.
```{r}
table(gtf.genes$gene_type[duplicated(gtf.genes$gene_name)])
```

Remove duplicated gene names which are mainly non-coding RNAs.
```{r}
removeDuplicated <- !duplicated(gtf.genes$gene_name)  # true when not duplicated
dge <- dge[removeDuplicated,,keep.lib.sizes=FALSE]
dim(dge)
```

Remove mitochondrial genes.  They are super abundant and can skew data.
```{r}
removeMT <-dge$genes$seqnames != "chrM" # true when not chrM
dge <- dge[removeMT,,keep.lib.sizes=FALSE]
dim(dge)
```

Make gene_name the row name.
```{r}
rownames(dge$counts) <- dge$genes$gene_name
```

Change sex column from numbers to letters.
```{r}
# male = 1, female = 2
sex_nums <- as.numeric(dge$samples$sex)
sex_letters <- gsub(1, "M", sex_nums)
sex_letters <- gsub(2, "F", sex_letters)
head(cbind(sex_nums, sex_letters)) # check
dge$samples$sex <- sex_letters
```

Subset by male or female if desired.
```{r}
keepMale <- dge$samples$sex == "M"
dge <- dge[,keepMale, keep.lib.sizes=FALSE]
dim(dge)
```

-----------------------------TTR TPM HISTOGRAM----------------------------------

Data is already in TPM (transcripts per million).
Create histogram of TPM TTR values.
```{r}
ttr.counts <- as.data.frame(dge$counts["TTR",])
colnames(ttr.counts) <- "ttr.counts"

log2.ttr.counts <- log2(ttr.counts + 0.01)
colnames(log2.ttr.counts) <- "log2.ttr.counts"
```

```{r}
h1 <- ggplot(ttr.counts, aes(x = ttr.counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = "A.", x=NULL, y=NULL) +
  xlab("TTR TPM") + ylab("# of Samples") + 
  geom_vline(xintercept = 20, col = "red") +
  theme_bw()

h2 <- ggplot(ttr.counts, aes(x = ttr.counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = "B.", x=NULL, y=NULL) +
  xlab("TTR TPM") + ylab("# of Samples") +
  xlim(0, 500) + geom_vline(xintercept = 20, col = "red") + 
  theme_bw() 

h3 <- ggplot(log2.ttr.counts, aes(x = log2.ttr.counts)) + 
  geom_histogram(bins = 100, fill = "gray", color = "black") + 
  labs(title = "GTEx\nMale hippocampus (n = 143)", x=NULL, y=NULL) +
  labs(x = expression(log[2](italic(TTR))), y = "# of samples") +
  theme_bw() +
  theme(plot.title = element_text(size=12)) +
  theme(axis.title.x = element_text(size = 12),
        axis.text.x = element_text(size = 12)) +
  theme(axis.title.y = element_text(size = 12),
        axis.text.y = element_text(size = 12)) +
  geom_vline(xintercept = log2(20) + 1, col = "blue") +
  geom_vline(xintercept = log2(20) - 1, col = "blue") +
  scale_y_continuous(breaks = c(1:9), limits = c(0,9)) + 
  scale_x_continuous(breaks = seq(-2,12, by = 2), limits = c(-4,13))
```

Arrange graphs in grid.
```{r}
plots1 <- list(h1,h2,h3)

layout1 <- rbind(c(1,2),c(3))

grid1 <- grid.arrange(grobs = plots1, layout_matrix = layout1)
```

------------------------------ASSIGN DIRTY/CLEAN/NEITHER------------------------

Lets looks at groups +/- 1 of the cutoff
```{r}
cutoff <- log2(20)
clean_threshold <- cutoff - 1 # values <= this are clean
clean_threshold
cutoff 
dirty_threshold <- cutoff + 1 # values >= this are dirty
dirty_threshold
```

Assign clean or dirty
```{r}
clean_or_dirty <- vector()
vector.log2.ttr.counts <- as.vector(log2.ttr.counts$log2.ttr.counts)

for (i in 1:length(vector.log2.ttr.counts)){
  if (vector.log2.ttr.counts[i] <= clean_threshold){
    clean_or_dirty <- c(clean_or_dirty, "clean")
  }
  else if (vector.log2.ttr.counts[i] >= dirty_threshold){
    clean_or_dirty <- c(clean_or_dirty, "dirty")
  }
  else {
    clean_or_dirty <- c(clean_or_dirty, "neither")
  }
}
table(clean_or_dirty)
```


```{r}
dge$samples$group <- factor(clean_or_dirty)
head(dge$samples)
```

```{r}
h4 <- h3 + 
  annotate("rect", 
              xmin = -Inf,
              xmax = clean_threshold, 
              ymin = 0, 
              ymax=Inf, 
              alpha=0.2, 
              fill="deepskyblue") +
  annotate("rect", 
              xmin = dirty_threshold,
              xmax = Inf, 
              ymin = 0, 
              ymax=Inf, 
              alpha=0.2, 
              fill="chocolate4")
h5 <- h1 + 
  annotate("rect", 
              xmin = -Inf,
              xmax = clean_threshold, 
              ymin = 0, 
              ymax=Inf, 
              alpha=0.2, 
              fill="deepskyblue") +
  annotate("rect", 
              xmin = dirty_threshold,
              xmax = Inf, 
              ymin = 0, 
              ymax=Inf, 
              alpha=0.2, 
              fill="chocolate4")
h4
h5
```

```{r}
setwd("~/TTR/03_GTEx/")

pdf(
  "Figures/GTEx_TTR_males_histogram.pdf",
  width = 4, height = 4)
h4
```

-----------------------FILTER LOWLY EXPRESSED GENES-----------------------------

TPM to CPM
```{r}
cpm <- cpm(dge)
```

Filter
```{r}
# produces a logical matrix
thresh <- cpm > 2 # cpm greater than 2

# keep genes that have at least 3 TRUES in each row of thresh
keep <- rowSums(thresh) >= 3
summary(keep)

# redefine object, filter by logical
dge.filtered <- dge[keep,,keep.lib.sizes = FALSE]

dim(dge)
dim(dge.filtered)
```
Went from 54,592 genes to 23,23419 genes.

--------------------------TRIMMED MEAN OF M-VALUES------------------------------

Now, we want to normalize gene expression distributions. We do this by normalizing 
library size differences in each sample. The method Trimmed Mean of M-values (TMM) 
is used with the calcNormFactors in the edgeR package. The normalization factors 
calculated here are used as a scaling factor for the library size.
  
Trimmed mean of M values (TMM) normalization estimates sequencing
depth after excluding genes for which the ratio of counts between a pair
of experiments is too extreme or for which the average expression is too
extreme. The edgeR software implements a TMM normalization (Biostar Handbook 2020).

```{r}
# creates another large DGEList 
dge.filtered.norm <- calcNormFactors(dge.filtered, method = "TMM")

# norm factor distribution
summary(dge.filtered.norm$samples$norm.factors)
```
Looking to be around ~1.

-----------------------------DESIGN MATRIX--------------------------------------

```{r}
age <- dge.filtered.norm$samples$age
```

Create a design matrix and specify the samples we want to compare to each other.
```{r}
# interaction computes a factor which represents the interaction of the given factors
group <- interaction(dge.filtered$samples$group)

design <- model.matrix(~ 0 + 
                         group +
                         age)
colnames(design) <- make.names(colnames(design))
colnames(design)[1:2] <- c("clean","dirty")
```

Run voom with quality weights. This combines observational-level weights with
  sample-specific quality weights in an experiment.
Normalizes expression intensities so that the log-ratios have similar distributions
  across a set of samples.
To quantile normalize, add normalize.method = "quantile".

Quality weights in RNA-seq data improves data quality when samples with high
  variance are present.  This variation is modeled by taking into account
  global intensity-dependent trends using voom.  
Sample-specific variability is taken into account with a log-linear model that 
  shares parameters between genes.
https://doi.org/10.1093/nar/gkv412


```{r}
v <- voomWithQualityWeights(dge.filtered.norm, design, plot = TRUE)
```

---------------------------------MDS PLOT---------------------------------------

```{r}
# blue is clean, brown is dirty
group_colors <- c("deepskyblue","chocolate4","gray")[v$targets$group]
# F or M for points
point_shapes <- v$targets$sex 
# Check everything matches
head(data.frame(v$targets$group, group_colors, v$targets$sex, point_shapes)) 
```


```{r}
#pdf("Figures/GTEx_TTR_males_MDS_voom.pdf", width = 4, height = 4)

mds <- plotMDS(
  v, # our data object
  top = 10, # only looking at top 10 genes
  pch = point_shapes,
  cex = 1, # point size
  dim.plot = c(1,2), # specifying principal components to be plotted
  plot = TRUE, # if TRUE then plot is created on current graphics device
  col = group_colors # assigning our colors for each genotype
)

legend(
  "topleft",
  pch = 16,
  legend = c("clean","dirty","neither"),
  col = c("deepskyblue","chocolate4","gray"),
  cex = 0.8
)

title("Top 10 Genes")
```

---------------------------------LINEAR MODELS----------------------------------

Fit linear models for comparisons of interest.
Linear modeling in limma is carried out using the lmFit and contrasts.fit
  functions.
  
What do lm.fit and contrasts.fit do?
  1) Fit a separate model to the expression values for each gene.
  2) An empirical Bayes moderation is carried out by borrowing info across all
     the gene to obtain more precise estimates of gene-wise variability.
  3) The model's residual variances are plotted against average expression 
     values in the next figure.

It can be seen from this plot that the variance is no longer dependent on the 
  mean expression level.
  
```{r}
# Fits linear model for each gene given a series of arrays
fit <- lmFit(
  v,  # object containing log-expression values for a series of arrays
  design  # design matrix
  )

```


Contrast design for differential expression.
```{r}
contrasts <-
  makeContrasts(
    dirty_vs_clean = dirty - clean,
    levels = colnames(design)
  )

head(contrasts)
```

Assign allComparisons to a vector for later use.
```{r}
allComparisons <- colnames(contrasts)
allComparisons # check
```

Run contrast analysis.
Given a linear model fit, compute estimated coefficients and standard errors
  for a given set of contrasts.
```{r}
vfit <- contrasts.fit(fit, contrasts = contrasts)
```

Compute differential expression based on the empirical Bayes moderation of the
  standard errors towards a common value.
```{r}
veBayesFit <- eBayes(vfit)
plotSA(veBayesFit, main = "Final Model: Mean-variance Trend")
```


decideTests() identifies which genes are significantly differentially expressed
  for each contrast from a fit object containing p-values and test statistics.
```{r}
sumTable <- 
  summary(decideTests(
    vfit,  # object
    # by default the method = "separate"
    adjust.method = "BH", 
    p.value = 0.05,
    lfc = 1  # numeric, minimum absolute log2-fold change required
  ))

head(sumTable)
```

--------------------------------------DEG TABLES--------------------------------

For each comparison...extract table of the top-ranked genes (DEGs) from a linear 
model fit AND output a table.
  
NOTE: Log2FC of 1 is equivalent to linear fold change of 2.

```{r}
setwd("~/TTR/03_GTEx/")

coef = 1
for (i in allComparisons) {
  
  # p < 1, log2fc > 0 ----------------------------------------------------------
  vTopTableAll <-
    topTable(
      veBayesFit, 
      coef = coef,  
      n = Inf, 
      p.value = 1,
      lfc = 0 
    )
  #output as txt file
  path <- paste("DEGs/DEGs_TTR_males_", i, "_FDRq1_Log2FC0.txt", sep = "") 
  write.table(
    vTopTableAll,
    path,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
  
  # p < 0.05, log2fc > 1 -------------------------------------------------------
  vTopTable1 <-
    topTable( 
      veBayesFit,  
      coef = coef,  
      n = Inf, 
      p.value = 0.05,
      lfc = 1
    )
  path <- paste("DEGs/DEGs_TTR_males_", i, "_FDRq0.05_Log2FC1.txt", sep = "") 
  write.table(
    vTopTable1,
    path,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )

  # increment the coefficient --------------------------------------------------
  coef <- coef + 1
}
```


----------------------------------VOLCANO PLOTS---------------------------------

Using the DEG tables we made, graph a volcano plot.
Read in file.
```{r}
setwd("~/TTR/03_GTEx/")

dirty_vs_clean <-
  read.table(
    "DEGs/DEGs_TTR_males_dirty_vs_clean_FDRq1_Log2FC0.txt",
    header = TRUE,
    sep = "\t",
    stringsAsFactors = F
  )
```

```{r}
color_values <- vector()
max <- nrow(dirty_vs_clean)

for(i in 1:max){
  if (dirty_vs_clean$adj.P.Val[i] < 0.05){
    if (dirty_vs_clean$logFC[i] > 1){
      color_values <- c(color_values, 1) # 1 when logFC > 1 and pval < 0.05
    }
    else if (dirty_vs_clean$logFC[i] < -1){
      color_values <- c(color_values, 2) # 2 when logFC < -1 and pval < 0.05
    }
    else{
      color_values <- c(color_values, 3) # 3 when -1 =< logFC <= 1 pval < 0.05
    }
  }
  else{
    color_values <- c(color_values, 3) # 3 when pval >= 0.05
  }
}

dirty_vs_clean$color_p0.05_lfc1 <- factor(color_values)
```

```{r}
hadjpval <- (-log10(max(dirty_vs_clean$adj.P.Val[dirty_vs_clean$adj.P.Val < 0.05], na.rm=TRUE)))
hadjpval
```

```{r}
p <-
  ggplot(data = dirty_vs_clean, 
         aes(x = logFC, # x-axis is logFC
             y = -log10(adj.P.Val), # y-axis will be log10 of adj.P.Val
             color = color_p0.05_lfc1)) + # color is based on factored Color column
  geom_point(size = 2) + # create scatterplot, alpha makes points transparent
  theme_bw() +
  theme(legend.position = "none") + # no legend
  scale_color_manual(values = c("red", "blue", "grey")) + # set factor colors
  labs(
    title = "", # no main title
    x = expression(log[2](FC)), # x-axis title
    y = expression(-log[10] ~ "(FDR adjusted " ~ italic("p") ~ "-value)") # y-axis title
  ) +
  scale_x_continuous(n.breaks = 6, limits = c(-3,7)) +
  scale_y_continuous(breaks = seq(0, 35, by = 10), limits = c(0,35)) +
  theme(plot.title = element_text(size=12)) +
  theme(axis.title.x = element_text(size = 12),
        axis.text.x = element_text(size = 12)) +
  theme(axis.title.y = element_text(size = 12),
        axis.text.y = element_text(size = 12)) +
  geom_hline(yintercept = hadjpval,
                     colour = "#000000",
                     linetype = "dashed") +
  geom_vline(xintercept = 1,
             colour = "#000000",
             linetype = "dashed") +
  geom_vline(xintercept = -1,
             colour = "#000000",
             linetype = "dashed") +
  ggtitle("GTEx\nMale hippocampus\nFDRq < 0.05, -1 > LogFC > 1") +
  geom_text_repel(data = subset(dirty_vs_clean, 
                                dirty_vs_clean$color_p0.05_lfc1 == 1),
                  aes(x = logFC, y= -log10(adj.P.Val), label = gene_name), 
                  color = "maroon",
                  fontface="italic",
                  max.overlaps = getOption("ggrepel.max.overlaps", 
                                           default = 50)) +
  geom_text_repel(data = subset(dirty_vs_clean,
                                dirty_vs_clean$color_p0.05_lfc1 == 2),
                  aes(x = logFC, y= -log10(adj.P.Val), label = gene_name),
                  color = "navyblue",
                  fontface="italic",
                  max.overlaps = getOption("ggrepel.max.overlaps",
                                           default = 25))
p
```

```{r}
setwd("~/TTR/03_GTEx/")
pdf("Figures/GTEx_TTR_males_volcano.pdf", width = 4, height = 4)
p
```

