---
title: "Hippocampus TPM Counts"
author: "Kennedi Todd"
date: "2/8/2021"
output:
  pdf_document: default
---
  Set working directory
```{r setup, include=FALSE, cache=FALSE}
#knitr::opts_knit$set(root.dir = "")
```

Load packages
```{r}
library(data.table)  # fread()
library(rtracklayer)  # to import gtf file
library(data.table)
library(edgeR)
library(IsoformSwitchAnalyzeR)
```

Get info for downloading files.
```{r}
data.files <- read.csv(file = "tbi_data_files.csv",header = TRUE, sep = ",")
donor.info <- read.csv(file = "DonorInformation.csv", header = TRUE, sep = ",")

# Subset hippocampus data
data.files <- data.files[data.files$structure_acronym == "HIP",]

# get donor metadata by joining tables based on donor_id
df.hip <- merge(data.files, donor.info, by = "donor_id")

# Exract info
links.hip <- df.hip$gene_level_fpkm_file_link
donorid.hip <- df.hip$donor_id
rnaid.hip <- df.hip$rnaseq_profile_id
donorsex.hip <- df.hip$sex
```

Output metadata file.
```{r}
write.table(df.hip, "hippocampus_metadata.txt", sep = "\t")
```

Get gene annotation file.
```{r}
gtf.file <- "rsem_GRCh38.p2.gtf"
gtf.gr <- rtracklayer::import(gtf.file)
gtf.df <- as.data.frame(gtf.gr)
```

Subset gtf.df to get gene list.  There are 50,283 observations.
```{r}
genes <- gtf.df[gtf.df$type == "gene",]
genes <- genes[,c(-8,-9)]
```

Download one file from links.hip to see the columns available to subset.
```{r}
url_path <- paste0("http://aging.brain-map.org", links.hip[1])
file <- fread(url_path, header = TRUE)
dim(file)
# We will want the gene_id column of at least one file.
# Merge the gene_id column with the gtf.df later
```
There are 50,283 observations.

Create gene info table.
```{r}
# The two columns we join by are different class types
# Make both gene_id columns the character type
file$gene_id <- as.character(file$gene_id)
gene.info <- as.data.frame(merge(file, genes, by = "gene_id"))
```

Check to see if our gene_id columns match.
We will save this data for all future downloaded files.
```{r}
all.equal(gene.info$gene_id, file$gene_id)
write.table(gene.info, "geneInfo.txt", sep = "\t")
```

Preview one file to see what we can subset.
```{r}
url_path <- paste0("http://aging.brain-map.org", links.hip[1])
file <- fread(url_path, header = TRUE)
```
We will subset the TPM column.

HIP
```{r}
counts1 <- data.frame(row.names = 1:50283)

for (i in 1:20) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts1 <- cbind(counts1, file)
  print(i)
  Sys.sleep(5)
}

counts2 <- data.frame(row.names = 1:50283)

for (i in 21:40) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts2 <- cbind(counts2, file)
  print(i)
  Sys.sleep(5)
}

counts3 <- data.frame(row.names = 1:50283)

for (i in 41:60) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts3 <- cbind(counts3, file)
  Sys.sleep(5)
}

counts4 <- data.frame(row.names = 1:50283)

for (i in 61:80) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts4 <- cbind(counts4, file)
  print(i)
  Sys.sleep(5)
}

counts5 <- data.frame(row.names = 1:50283)

for (i in 81:94) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts5 <- cbind(counts5, file)
  print(i)
  Sys.sleep(5)
}

counts <- cbind(counts1, counts2, counts3, counts4, counts5)
colnames(counts) <- c(donorid.hip)
rownames(counts) <- gene.info$gene_id

sex.info <- as.data.frame(cbind(donorid.hip,donorsex.hip))
colnames(sex.info) <- c("donor_id","sex")

write.table(counts,"hippocampus_tpm_counts.txt", sep = "\t")
write.table(sex.info, "hippocampus_donorid_sex.txt", sep = "\t")
```


Get info for downloading files.
```{r}
data.files <- read.csv(file = "tbi_data_files.csv",header = TRUE, sep = ",")
donor.info <- read.csv(file = "DonorInformation.csv", header = TRUE, sep = ",")

# Extract sample info
tissue_acronyms <- unique(data.files$structure_acronym)
tissue_acronyms

# Subset FWM data
data.files.FWM <- subset(data.files, structure_acronym == "FWM")

# we can get donor sex by joining tables based on donor_id
df.FWM <- merge(data.files.FWM, donor.info, by = "donor_id")

# Exract info
links.FWM <- df.FWM$gene_level_fpkm_file_link
donorid.FWM <- df.FWM$donor_id
rnaid.FWM <- df.FWM$rnaseq_profile_id
donorsex.FWM <- df.FWM$sex
```

FWM
```{r}
counts1 <- data.frame(row.names = 1:50283)

for (i in 1:20) {
  url_path <- paste0("http://aging.brain-map.org", links.FWM[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts1 <- cbind(counts1, file)
  print(i)
  Sys.sleep(5)
}

counts2 <- data.frame(row.names = 1:50283)

for (i in 21:40) {
  url_path <- paste0("http://aging.brain-map.org", links.FWM[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts2 <- cbind(counts2, file)
  print(i)
  Sys.sleep(5)
}

counts3 <- data.frame(row.names = 1:50283)

for (i in 41:60) {
  url_path <- paste0("http://aging.brain-map.org", links.FWM[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts3 <- cbind(counts3, file)
  Sys.sleep(5)
}

counts4 <- data.frame(row.names = 1:50283)

for (i in 61:80) {
  url_path <- paste0("http://aging.brain-map.org", links.FWM[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts4 <- cbind(counts4, file)
  print(i)
  Sys.sleep(5)
}

counts5 <- data.frame(row.names = 1:50283)
# only 92 for FWM 
for (i in 81:93) {
  url_path <- paste0("http://aging.brain-map.org", links.FWM[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts5 <- cbind(counts5, file)
  print(i)
  Sys.sleep(5)
}

counts <- cbind(counts1, counts2, counts3, counts4, counts5)
colnames(counts) <- c(donorid.FWM)
rownames(counts) <- gene.info$gene_id

sex.info <- as.data.frame(cbind(donorid.FWM,donorsex.FWM))
colnames(sex.info) <- c("donor_id","sex")

write.table(counts,"FWM_tpm_counts.txt", sep = "\t")
write.table(sex.info, "FWM_donorid_sex.txt", sep = "\t")
```


Get info for downloading files.
```{r}
data.files <- read.csv(file = "tbi_data_files.csv",header = TRUE, sep = ",")
donor.info <- read.csv(file = "DonorInformation.csv", header = TRUE, sep = ",")

# Extract sample info
tissue_acronyms <- unique(data.files$structure_acronym)
tissue_acronyms

# Subset PCx data
data.files.PCx <- subset(data.files, structure_acronym == "PCx")

# we can get donor sex by joining tables based on donor_id
df.PCx <- merge(data.files.PCx, donor.info, by = "donor_id")

# Exract info
links.PCx <- df.PCx$gene_level_fpkm_file_link
donorid.PCx <- df.PCx$donor_id
rnaid.PCx <- df.PCx$rnaseq_profile_id
donorsex.PCx <- df.PCx$sex
```

PCx
```{r}
counts1 <- data.frame(row.names = 1:50283)

for (i in 1:8) {
  url_path <- paste0("http://aging.brain-map.org", links.PCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts1 <- cbind(counts1, file)
  print(i)
  Sys.sleep(5)
}

counts1.2 <- data.frame(row.names = 1:50283)

for (i in 9:20) {
  url_path <- paste0("http://aging.brain-map.org", links.PCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts1 <- cbind(counts1, file)
  print(i)
  Sys.sleep(5)
}

counts2 <- data.frame(row.names = 1:50283)

for (i in 21:40) {
  url_path <- paste0("http://aging.brain-map.org", links.PCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts2 <- cbind(counts2, file)
  print(i)
  Sys.sleep(5)
}

counts3 <- data.frame(row.names = 1:50283)

for (i in 41:60) {
  url_path <- paste0("http://aging.brain-map.org", links.PCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts3 <- cbind(counts3, file)
  Sys.sleep(5)
}

counts4 <- data.frame(row.names = 1:50283)

for (i in 61:80) {
  url_path <- paste0("http://aging.brain-map.org", links.PCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts4 <- cbind(counts4, file)
  print(i)
  Sys.sleep(5)
}

counts5 <- data.frame(row.names = 1:50283)

for (i in 81:91) {
  url_path <- paste0("http://aging.brain-map.org", links.PCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts5 <- cbind(counts5, file)
  print(i)
  Sys.sleep(5)
}

counts <- cbind(counts1, counts2, counts3, counts4, counts5)
colnames(counts) <- c(donorid.PCx)
rownames(counts) <- gene.info$gene_id

sex.info <- as.data.frame(cbind(donorid.PCx,donorsex.PCx))
colnames(sex.info) <- c("donor_id","sex")

write.table(counts,"PCx_tpm_counts.txt", sep = "\t")
write.table(sex.info, "PCx_donorid_sex.txt", sep = "\t")
```

```{r}
data.files <- read.csv(file = "tbi_data_files.csv",header = TRUE, sep = ",")
donor.info <- read.csv(file = "DonorInformation.csv", header = TRUE, sep = ",")

# Extract sample info
tissue_acronyms <- unique(data.files$structure_acronym)
tissue_acronyms

# Subset TCx data
data.files.TCx <- subset(data.files, structure_acronym == "TCx")

# we can get donor sex by joining tables based on donor_id
df.TCx <- merge(data.files.TCx, donor.info, by = "donor_id")

# Exract info
links.TCx <- df.TCx$gene_level_fpkm_file_link
donorid.TCx <- df.TCx$donor_id
rnaid.TCx <- df.TCx$rnaseq_profile_id
donorsex.TCx <- df.TCx$sex
```
TCx
```{r}
counts1 <- data.frame(row.names = 1:50283)

for (i in 1:20) {
  url_path <- paste0("http://aging.brain-map.org", links.TCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts1 <- cbind(counts1, file)
  print(i)
  Sys.sleep(5)
}
counts2 <- data.frame(row.names = 1:50283)

for (i in 21:30) {
  url_path <- paste0("http://aging.brain-map.org", links.TCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts2 <- cbind(counts2, file)
  print(i)
  Sys.sleep(5)
}

for (i in 31:40) {
  url_path <- paste0("http://aging.brain-map.org", links.TCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts2 <- cbind(counts2, file)
  print(i)
  Sys.sleep(5)
}

counts3 <- data.frame(row.names = 1:50283)

for (i in 41:60) {
  url_path <- paste0("http://aging.brain-map.org", links.TCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts3 <- cbind(counts3, file)
  Sys.sleep(5)
}

counts4 <- data.frame(row.names = 1:50283)

for (i in 61:80) {
  url_path <- paste0("http://aging.brain-map.org", links.TCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts4 <- cbind(counts4, file)
  print(i)
  Sys.sleep(5)
}

counts5 <- data.frame(row.names = 1:50283)

for (i in 81:94) {
  url_path <- paste0("http://aging.brain-map.org", links.TCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts5 <- cbind(counts5, file)
  print(i)
  Sys.sleep(5)
}

for (i in 95:98) {
  url_path <- paste0("http://aging.brain-map.org", links.TCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts5 <- cbind(counts5, file)
  print(i)
  Sys.sleep(5)
}

for (i in 99:99) {
  url_path <- paste0("http://aging.brain-map.org", links.TCx[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts5 <- cbind(counts5, file)
  print(i)
  Sys.sleep(5)
}

counts <- cbind(counts1, counts2, counts3, counts4, counts5)
colnames(counts) <- c(donorid.TCx)
rownames(counts) <- gene.info$gene_id

sex.info <- as.data.frame(cbind(donorid.TCx,donorsex.TCx))
colnames(sex.info) <- c("donor_id","sex")

write.table(counts,"TCx_tpm_counts.txt", sep = "\t")
write.table(sex.info, "TCx_donorid_sex.txt", sep = "\t")
```

