---
title: "Get Hippocampus Data: TPM Counts"
author: "Kennedi Todd"
date: "2/8/2021"
output: html_document
---

```{r}
setwd("/Users/m214960/Documents/R_Workspace/Allen_Brain_Atlas/raw_data")
```

```{r}
library(data.table)
library(edgeR)
library(rtracklayer)
```

Get info for downloading files.
```{r}
data.files <- read.csv(file = "tbi_data_files.csv",header = TRUE, sep = ",")
donor.info <- read.csv(file = "DonorInformation.csv", header = TRUE, sep = ",")

# Extract sample info
tissue_acronyms <- unique(data.files$structure_acronym)
tissue_acronyms

# Subset hippocampus data
data.files.hip <- subset(data.files, structure_acronym == "HIP")

# we can get donor sex by joining tables based on donor_id
df.hip <- merge(data.files.hip, donor.info, by = "donor_id")

# Exract info
links.hip <- df.hip$gene_level_fpkm_file_link
donorid.hip <- df.hip$donor_id
rnaid.hip <- df.hip$rnaseq_profile_id
donorsex.hip <- df.hip$sex
```

Get gene annotation file.
```{r}
gtf.file <- "rsem_GRCh38.p2.gtf"
gtf.gr <- rtracklayer::import(gtf.file)
gtf.df <- as.data.frame(gtf.gr)
```

Subset gtf.df to get gene list.  There are 50,283 observations.
```{r}
genes <- subset(gtf.df, type == "gene")
```

Download one file from links.hip to see the columns that are available to subset.
There are 50,283 observations.
```{r}
url_path <- paste0("http://aging.brain-map.org", links.hip[1])
file <- fread(url_path, header = TRUE)
# We will want the gene_id column of at least one file.
# Merge the gene_id column with the gtf.df later
```

Create gene info table.
```{r}
# The two columns we join by are different class types
# Make both gene_id columns the character type
file$gene_id <- as.character(file$gene_id)
gene.info <- as.data.frame(merge(file, genes, by = "gene_id"))
```

Reformat the table to exclude file specific values.
Keep gene info.
```{r}
keep.columns <- c(1,3:5,8:17)
gene.info <- gene.info[,keep.columns]
```


Check to see if our gene_id columns match.
```{r}
all.equal(gene.info$gene_id, file$gene_id)
```


Output files.
```{r}
write.table(gene.info, "gene_info_for_counts_file.txt", sep = "\t")
```


Preview one file to see what we can subset.
```{r}
url_path <- paste0("http://aging.brain-map.org", links.hip[1])
file <- fread(url_path, header = TRUE)
```


```{r}
counts1 <- data.frame(row.names = 1:50283)

for (i in 1:20) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts1 <- cbind(counts1, file)
  print(i)
  Sys.sleep(5)
}
```

```{r}
counts2 <- data.frame(row.names = 1:50283)

for (i in 21:40) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts2 <- cbind(counts2, file)
  print(i)
  Sys.sleep(5)
}
```

```{r}
counts3 <- data.frame(row.names = 1:50283)

for (i in 41:60) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts3 <- cbind(counts3, file)
  Sys.sleep(5)
}
```

```{r}
counts4 <- data.frame(row.names = 1:50283)

for (i in 61:80) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts4 <- cbind(counts4, file)
  print(i)
  Sys.sleep(5)
}
```

```{r}
counts5 <- data.frame(row.names = 1:50283)

for (i in 81:94) {
  url_path <- paste0("http://aging.brain-map.org", links.hip[i])
  file <- fread(url_path, header = TRUE, select = "TPM")
  file <- as.vector(file)
  counts5 <- cbind(counts5, file)
  print(i)
  Sys.sleep(5)
}
```

Create final counts file.
```{r}
counts <- cbind(counts1, counts2, counts3, counts4, counts5)
colnames(counts) <- c(donorid.hip)
rownames(counts) <- gene.info$gene_id

sex.info <- as.data.frame(cbind(donorid.hip,donorsex.hip))
colnames(sex.info) <- c("donor_id","sex")

write.table(counts,"hippocampus_tpm_counts.txt", sep = "\t")
write.table(sex.info, "metadata.txt", sep = "\t")
```







